\chapter{Introduction}

The basic question of computational complexity---``how hard is this problem for
a computer to solve?''---is central to nearly every topic in computer science.
And yet the formalisms of complexity theory often seem, in my own experience,
intimidatingly abstract, phrased in terms of intangible models of computation
such as non-deterministic Turing machines and oracles.

The remedy, I believe, lies in studying complexity theory through the lens of
\emph{puzzles} and \emph{games}.  Not only do they provide a concrete grounding
for the abstractions, they also offer a particularly insightful, accessible,
and most importantly fun approach to understanding complexity theory.  In fact,
many of the most popularly known and appreciated results in complexity theory
are those about so-called ``\NP-complete puzzles'', such as Sudoku, and
``\PSPACE-complete games'', such as Checkers and Go.

This thesis emphasizes that approach in its exploration of a particularly
foundational, yet often overlooked, ladder of complexity classes known as the
\emph{polynomial hierarchy}.  \NP{} is the class of (one-player) ``puzzles'',
and \PSPACE{} is the class of (two-player) ``games'' of polynomial length; the
polynomial hierarchy, then, lies in the middle, encompassing games of
\emph{fixed} length.  Through this lens, the (in)famous \P-vs-\NP{} question is
but the first in a ladder of questions that are, arguably, just as crucial and
impactful.

%The polynomial hierarchy is as central to
%complexity theory as the \P-vs-\NP{} problem is well-known.

\section{Overview}

This document is structured as follows.  First, \cref{ch:background,ch:boolean}
establish preliminary background concepts and conventions adopted throughout
this thesis.  Next, \cref{ch:circuit} lays the central theoretical groundwork,
defining the \emph{polynomial hierarchy} through a fundamental family of
problems known as the \Problem{Circuit Satisfiability} games.  Next,
\cref{ch:misc} explores a novel family of games generalized from the
\Problem{Graph 3-Colorability} puzzle and establishes \emph{hardness} bounds on
each of those games.  Finally, \cref{ch:conclusion} concludes by discussing the
future directions of this work and its broader implications.

\section{Prior work and inspirations}

Much of the background exposition on complexity theory referenced in this thesis
is reproduced from Christos Papadimitriou's textbook,
\citet{papadimitriou.cc} (though many of the foundational ideas were
originally introduced/proven elsewhere, e.g.
\citet{cook.np,levin.np,stockmeyer.ph}), reframed through the
puzzles-and-games perspective and supplemented with a few comments on intuition.

The main family of games explored in this thesis, fixed-turn
\Problem{3-Colorability} games (\cref{ch:misc}), is a generalization of
(one-turn) \Problem{3-Colorability}, a well-known \NP-complete puzzle originally
proven \NP-complete by \citet{karp.np}.  Others have studied (multi-turn)
game generalizations of \Problem{3-Colorability}, but all versions that I've
encountered are \PSPACE-complete, in which the number of turns played during the
game scales proportionally with the size of the graph
\citep{bodlaender.coloring,bh.placement,kbd.impartial,cpss.coloring,schaefer.games}. As far as I'm aware, the variations I explore here—with fixed
numbers of turns regardless of the size of the graph—is unexplored, and the main
theorem about its \SigmaP k-completeness (\cref{th:yayay}) is novel.  The basic
idea underlying my proof is the composition of two well-known results:
\begin{itemize}[nosep]
  \item \citet{karp.np}'s classic proof of the \NP-hardness of the \Problem{3-Colorability} puzzle, via a reduction from \Problem{3CNF-Satisfiability};
  \item 's transformation from boolean circuits to equivalent 3CNF-clauses.
\end{itemize}

Without further ado, let's begin.


%TODO: outline/overview of chapters, after those chapters are written

%TODO: also give general citations here, e.g. papadimitriou for many
%foundational background info, etc.
%
%TODO: notation table also belongs in this chapter i think

%p-vs-np well known, polynomial-hierarchy central

%puzzles and games; hierarchy lies in the interstices.  we examine a few
%interesting (by no means exhaustive, or even close to comprehensive)
%np-complete puzzles with pspace-complete analogues, and we









%Famously central to the theory of computational complexity is the \P-vs-\NP{}
%question, and essential to our understanding of that question is the study of
%\NP-complete problems such as the Boolean Satisfiability puzzle, the Graph
%Colorability puzzle, and countless more.  Puzzles like these, which nearly any
%layperson can appreciate, offer a particularly insightful, intuitive, and
%\emph{fun} lens through which to study computational complexity. Explorations
%of more complex problem-classes such as \PSPACE{} can be similarly approached
%through the study of strategic decision \emph{games} such as Othello, Checkers,
%and Go.
%
%What lies in the interstices between \emph{puzzles} and \emph{games}?  How do
%we take a puzzle and generalize it into a game, and what are the puzzle-games
%we encounter along the way?  And how hard exactly are these puzzle-games to
%decide?  These questions are the focus of my thesis.

%So far, I have explored these questions from three angles:
%\begin{enumerate}
%
%  \item \label{itm:intro.q.generation} Puzzle generation.  If I wish to solve a
%    puzzle, you can play a game with me by constructing puzzle \emph{instances}
%    for me to solve.  For instance, \emph{solving} Sudoku is an \NP-complete
%    problem; your task is to \emph{generate} (partially-filled) Sudoku boards
%    for me to solve.
%
%    How hard is it to do so?  Moreover, how hard is it to generate \emph{good}
%    puzzle instances, for various definitions of \emph{good} (sufficiently
%    challenging to solve, or having unique solutions, or solvable/unsolvable by
%    certain strategies)?
%
%    % lauren sanchis
%
%  \item \label{itm:intro.q.pspace} \PSPACE-complete games derived from
%    \NP-complete puzzles.  A canonical \NP-complete puzzle is the \Problem{sat}
%    (Boolean Satisfiability) puzzle: given a Boolean formula \(\phi(x_1, \dots,
%    x_n)\), does there exist an assignment to its inputs \(x_1, \dots, x_n\)
%    such that \(\phi(\dots) = 1\)? In an analogous game, two players alternate
%    turns assigning \(x_1, \dots, x_n\); player 1 wins if \(\phi(\dots)=1\),
%    and player 2 wins if \(\phi(\dots)=0\).  Does either player have a
%    (guaranteed) winning strategy?  This game, known as \Problem{qsat}
%    (Quantified Satisfiability), is a canonical example of a \PSPACE-complete
%    game.
%
%    Can other \NP-complete puzzles be similarly generalized into games?  Will
%    those games also be \PSPACE-complete?
%
%    % schaefer
%
%  \item \label{itm:intro.q.ph} Fixed-turn games and the polynomial hierarchy.
%    In between the complexity classes \NP{} and \PSPACE{} lies a chain of
%    increasingly-complex problem-classes known as the \emph{polynomial
%    hierarchy}.  In some cases, problems in the polynomial hierarchy may be
%    thought of as game generalizations of \NP-complete puzzles with a
%    \emph{fixed} number of turns.  For instance, in a two-turn version of
%    \Problem{sat}, inputs are partitioned into two (disjoint) groups \(X_1\)
%    and \(X_2\); on turn 1, player 1 assigns \(X_1\), and on turn 2, player 2
%    assigns \(X_2\).  As before, player 1 (respectively 2) wins if
%    \(\phi(\dots) = 1\) (respectively \(0\)).  Determining whether player 1 has
%    a winning strategy is complete for a complexity class known as
%    \(\SigmaP2\), which lies just above \NP{} in the hierarchy, and analogous
%    games with \(k\) turns are \(\SigmaP k\)-complete.
%
%    Do polynomial-hierarchy generalizations of other \NP-complete puzzles
%    exist?
%
%\end{enumerate}
%
%\Cref{ch:progress} discusses my progress so far in each of these areas.
%Questions \ref{itm:intro.q.generation} and \ref{itm:intro.q.pspace} have been
%explored in-depth by others, while question \ref{itm:intro.q.ph} appears to be
%scarcely explored.  As such, I provide only brief summaries of/reflections on
%the existing work pertaining to \ref{itm:intro.q.generation} and
%\ref{itm:intro.q.pspace}.  Meanwhile, I describe in greater detail question
%\ref{itm:intro.q.ph}, which is the focus of my explorations so far.
%
%\Cref{ch:future} summarizes the primary questions \& goals that will guide
%my exploration next semester.
%
%Finally, \cref{ch:bib} contains an annotated bibliography of existing work
%pertaining to each of these topics.


