\documentclass{final-report}

\title{Games for one, games for two!}

\subtitle{computationally complex fun for polynomial-hierarchical families}

\author{Kye Shi}
\advisor{Nicholas Pippenger}
\reader{Arthur T. Benjamin}
\date{May 2022}

%\titlespacing\chapter{0em}{0em}{0em}
%\setlength{\beforechapskip}{0em}

\usepackage{caption}
\captionsetup{labelsep=period, labelfont=bf, margin=.5in}

\usepackage{tocloft}
\setlength\cftbeforetoctitleskip{-.5in}

%\usepackage{natbib}
%\bibliography{bibliography.bib}
\bibliography{newbib.bib}

\tikzset{
  gates/.style={
    row sep=1em/2,
    column sep=2em,
    matrix of nodes,
  },
  every circuit symbol/.style={
    fill,
    fill opacity=1/8,
    anchor=output,
  },
  %vertex/.style={
  %  circle,
  %  draw,
  %  minimum size=1em/2,
  %  inner sep=0pt,
  %},
  %vertex label/.style={
  %  text height=2em/3,
  %  text depth=1em/3,
  %},
  %wire/.style={
  %  rounded corners=1em/4,
  %  to path={
  %    -- ($ (\tikztostart)!1em!(\tikztostart -| \tikztotarget) $)
  %    -- ($ (\tikztotarget)!1em!(\tikztostart |- \tikztotarget) $)
  %    -- (\tikztotarget)
  %  },
  %},
}

\begin{document}

\pagenumbering{roman}
\frontmatter
\maketitle

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

%perspective of viewing

In the first half of this thesis, we explore the \emph{polynomial-time
hierarchy}, emphasizing an intuitive perspective that associates decision
problems in the polynomial hierarchy to combinatorial games with fixed numbers
of turns. Specifically, problems in \P{} are thought of as \(0\)-turn games,
\NP{} as \(1\)-turn ``puzzle'' games, and in general \(\SigmaP k\) as \(k\)-turn
games, in which decision problems answer the binary question, ``can the starting
player guarantee a win?'' We introduce the formalisms of the polynomial
hierarchy through this perspective, alongside definitions of \(k\)-turn
\Problem{Circuit Satisfiability} games, whose \SigmaP k-completeness is assumed
from prior work (we briefly justify this assumption on intuitive grounds, but no
proof is given).

In the second half, we introduce and explore the properties of a novel family of
games called the \(k\)-turn \Problem{Graph 3-Colorability} games.  By embedding
boolean circuits in proper graph 3-colorings, we construct reductions from
\(k\)-turn \Problem{Circuit Satisfiability} games to \(k\)-turn
\Problem{3-Colorability} games, thereby showing that \(k\)-turn
\Problem{3-Colorability} is \SigmaP k-complete.

%Viewing the polynomial hierarchy as classes of fixed-turn-games frames the
%infamous \P-vs-\NP{} question as part of a more general question, ``does the
%number of turns in a game make a difference (to its difficulty)?''  Given how
%thoroughly this question has been explored at the first level,

Finally, we conclude by discussing possible future generalizations of this work,
vis-\`a-vis extending arbitrary \NP-complete puzzles to interesting \SigmaP
k-complete games.

%Ultimately, I hope that this games-and-puzzles interpretation of the polynomial
%hierarchy sheds intuitive light on the (sometimes daunting) abstractions of
%complexity theory,




%In this thesis, I give a detailed exposition of the \emph{polynomial-time
%hierarchy}, particularly emphasizing the perspective
%
%lens of viewing
%
%combinatorial games with fixed numbers of turns.

\pagestyle{plain}
%\vspace*{-1in}
\newpage
\tableofcontents

\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}

First and foremost, I wish to thank my thesis advisor, Professor Nicholas
Pippenger, for generously signing on to advise me despite being retired, for
being tremendously patient with me despite how frequently I showed up to our
weekly meetings empty-handed, and of course for his helpful guidance throughout
this project.  While we're at it, let me also thank all the other wonderful
professors who have mentored me through the years—Professor Stephan Garcia,
Professor Dagan Karp, Professor Lucas Bang, Professor Weiqing Gu—you helped make
me the mathematician and the person I am now.

I am also grateful to my friends at
\href{http://mathily.org}{MathILy}—sarah-marie, Tom, Hannah, Brian, Corrine,
Max, Josh—you're the reason I fell in love with math in the first place, and
working with you continually inspires me to approach math with unrelenting
levity.

%Thank you, especially, to Corrine and Josh, for being there to support
%me that one time I \emph{really} needed it—

And finally, dearest thanks to my close friends outside of math: Sophia Cheng,
for supporting me in every way and for being a fabulous dance partner; Forest
Kobayashi, for befriending me ; Cole Kurashige, for his
sagely life advice; Kaveh Pezeshki, my OneWheel buddy for life; and last but not
least, my mom, for funding my exorbitant college education (and also, of course,
for taking care of me my entire life).  Cheers!

\cleardoublepage
\pagenumbering{arabic}

\pagestyle{thesis}

%draft outline:
%
%\begin{itemize}
%  \item Introduction and overview.
%  \item Basic complexity background: P, reductions, completeness
%  \item More complexity background---NP, PSPACE, polynomial hierarchy as games,
%    etc. Haven't made up my mind yet about whether to introduce this abstractly
%    on its own, or introduce it concurrently with CSAT games to keep things
%    concrete/tangible.  I'm leaning the latter.
%    \begin{itemize}
%      \item Circuit value, satisfiability games
%    \end{itemize}
%  \item SAT variation games
%    \begin{itemize}
%      \item Circuits could just as well be plain boolean formulae, or CNF
%        formulae, etc.; list some prior work on other known SAT games in
%        PSPACE, polynomial-hierarchy, etc.  (paper on this in ann. bib.)
%    \end{itemize}
%  \item Graph coloring games
%    \begin{itemize}
%      \item Intro---overview of existing results: NP-completeness of graph
%        coloring, col and snort as PSPACE-complete games.
%      \item Custom presentation of NP-completeness proof via circuit reduction
%      \item Introduction of 2-turn game, 3-turn game, k-turn game, etc.
%      \item Complexity claims and proofs
%    \end{itemize}
%  \item Exact-covering games
%  \item three-dimensional matching games
%  \item Conclusion % only a few of countlessly many problems introduced/explored
%\end{itemize}

%TODO: table of contents for figures, or theorems, definitions, etc.
%
%TODO: at some point i also need to figure out whether/how to make a glossary/index


\mainmatter

\include{chapter/intro}
\include{chapter/background}
\include{chapter/boolean}
\include{chapter/games}
\include{chapter/misc}
\include{chapter/conclusion}

%\include{chapter/intro}
%\include{chapter/progress}
%\include{chapter/future}
%\include{chapter/bib}

%\bibliographystyle{hmcmath}
%\bibliography{newbib.bib}
%\bibliography{shitbib.bib}

\printbibliography[heading=bibintoc]

\end{document}
